---
title: "R Project: Online Course Engagement Data Analysis"
author: "Mallikarjuna Mannem"
date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
output:
  rmdformats::readthedown:
    css: gvsu_readthedown_theme.css
---


## Objectives

Describe probability as a foundation of statistical modeling, including inference and maximum likelihood estimation.

Determine and apply the appropriate generalized linear model for a specific data context.

Conduct model selection for a set of candidate models.

Communicate the results of statistical models to a general audience.

## About Dataset

**Description:**

This dataset captures user engagement metrics from an online course platform, facilitating analyses on factors influencing course completion. It includes user demographics, course-specific data, and engagement metrics.

*Source:* https://www.kaggle.com/datasets/rabieelkharoua/predict-online-course-engagement-dataset


```{r, warning=FALSE}
# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(caret)
library(reshape2)
library(pROC)
```


### Features

**UserID**: Unique identifier for each user

**CourseCategory**: Category of the course taken by the user (e.g., Programming, Business, Arts)

**TimeSpentOnCourse**: Total time spent by the user on the course in hours

**NumberOfVideosWatched**: Total number of videos watched by the user

**NumberOfQuizzesTaken**: Total number of quizzes taken by the user

**QuizScores**: Average scores achieved by the user in quizzes (percentage)

**CompletionRate**: Percentage of course content completed by the user

**DeviceType**: Type of device used by the user (Device Type: Desktop (0) or Mobile (1))

**CourseCompletion** (Target Variable): Course completion status (0: Not Completed, 1: Completed)

```{r}
# Load the data
data <- read.csv("online_course_engagement_data.csv")

# View the structure of the dataset
str(data)

```

```{r}
# Summary of the data
summary(data)

```

```{r}
# Check for missing values
sum(is.na(data))
```


## Descriptive Statistics and Probability
To describe probability as a foundation of statistical modeling, we will perform some basic descriptive statistics and visualize the distribution of key features.

```{r}
# Descriptive statistics
summary(data)

# Visualize the distribution of time spent on course
ggplot(data, aes(x = TimeSpentOnCourse)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Distribution of Time Spent on Course", x = "Time Spent (hours)", y = "Frequency")

# Visualize the distribution of quiz scores
ggplot(data, aes(x = QuizScores)) +
  geom_histogram(binwidth = 5, fill = "green", color = "black") +
  labs(title = "Distribution of Quiz Scores", x = "Quiz Scores (%)", y = "Frequency")

# Visualize the distribution of course completion status
ggplot(data, aes(x = CourseCompletion)) +
  geom_bar(fill = "purple", color = "black") +
  labs(title = "Distribution of Course Completion Status", x = "Course Completion (0: Not Completed, 1: Completed)", y = "Count")


# Course completion by course category (count)
ggplot(data, aes(x = CourseCategory, fill = CourseCompletion)) +
  geom_bar(position = "dodge") +
  labs(title = "Course Completion by Course Category", x = "Course Category", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Generalized Linear Model (GLM)
Next, we determine and apply the appropriate generalized linear model for predicting course completion status.

```{r}
# Convert CourseCompletion to a factor
data$CourseCompletion <- as.factor(data$CourseCompletion)

# Fit a logistic regression model
glm_model <- glm(CourseCompletion ~ TimeSpentOnCourse + NumberOfVideosWatched + NumberOfQuizzesTaken + QuizScores + CompletionRate + DeviceType, 
                 data = data, 
                 family = binomial)

# Summary of the model
summary(glm_model)
```
The reduction in deviance from the null model to the residual model indicates a good fit. The AIC of the selected model is lower than that of the initial model, indicating improved performance.

The stepwise selection process identified key predictors: *TimeSpentOnCourse*, *NumberOfVideosWatched*, *NumberOfQuizzesTaken*, *QuizScores*, and *CompletionRate.* These predictors significantly influence course completion, providing a well-balanced model with good fit and interpretability. This final model effectively predicts course completion while avoiding overfitting.

## Model Selection
We will conduct model selection using criteria such as AIC and BIC.

Model selection helps identify the best model that balances complexity and performance. We used stepwise selection based on AIC (Akaike Information Criterion) for this purpose.

**AIC:** A measure of model quality that balances goodness of fit and complexity. Lower AIC values indicate a better model.

**Stepwise Selection:** Iteratively adds/removes predictors to minimize AIC, stopping when no further improvement is possible.

```{r}
# Model selection using AIC
selected_model <- step(glm_model, direction = "both", trace = FALSE)

# Compare models using AIC and BIC
library(MuMIn)
model_comparison <- model.sel(glm_model, selected_model)
model_comparison

```

```{r}
summary(selected_model)

```


## Model Assessment and Visualization
Assess the fitted model and visualize the results to communicate findings to a general audience.

```{r}
# Predict probabilities of course completion
data$PredictedProb <- predict(glm_model, type = "response")

# Create a confusion matrix
table(data$CourseCompletion, data$PredictedProb > 0.5)

# Visualize predicted probabilities
ggplot(data, aes(x = PredictedProb, fill = CourseCompletion)) +
  geom_histogram(binwidth = 0.1, position = "dodge", color = "white") +
  labs(title = "Predicted Probabilities of Course Completion", x = "Predicted Probability", y = "Count")

```


## ROC Curve

The Receiver Operating Characteristic (ROC) curve is a graphical representation of the performance of a binary classification model. It plots the True Positive Rate (TPR) against the False Positive Rate (FPR) at various threshold settings.


**True Positive Rate (TPR) / Sensitivity / Recall:** This is the proportion of actual positives (students who completed the course) that are correctly identified by the model. It is plotted on the Y-axis.

**False Positive Rate (FPR):** This is the proportion of actual negatives (students who did not complete the course) that are incorrectly identified as positives by the model. It is plotted on the X-axis.

**Thresholds:** Different points on the ROC curve correspond to different thresholds used to classify whether a student will complete the course or not. By varying the threshold, you can trade off between TPR and FPR.

**Area Under the Curve (AUC):** This is a single scalar value summarizing the overall performance of the classifier. The AUC ranges from 0 to 1, where:

AUC = 0.5: The model has no discrimination capability, equivalent to random guessing.

AUC > 0.5: The model has some discrimination capability, with higher values indicating better performance.

AUC = 1: The model has perfect discrimination capability.


**Shape of the ROC Curve:** The closer the ROC curve follows the top-left corner (i.e., higher TPR and lower FPR), the better the model performs.

**AUC Value:** The AUC value provides an aggregate measure of the model's performance. For example, an AUC of 0.85 would indicate that the model has a good performance, meaning it has a high probability of correctly distinguishing between students who will complete the course and those who will not.

In the context, the ROC curve and AUC help evaluate how well the *logistic regression model* distinguishes between students who complete the course and those who do not. Let's interpret the ROC curve and AUC value from the provided code.

```{r}

# Plot ROC curve
roc_curve <- roc(data$CourseCompletion, data$PredictedProb)
plot(roc_curve, main = "ROC Curve for Course Completion Prediction")

# Calculate the Area Under the Curve (AUC)
auc_value <- auc(roc_curve)
auc_value
```



## Communicating Results

**Insights from Visualizations**:

**Time Spent on Course and Quiz Scores:** Significant predictors of course completion, as more time spent and higher quiz scores are associated with higher completion rates.

**Course Category:** Completion rates vary across different course categories, indicating that course type might influence completion rates.


### Summary and Interpretation:

**Descriptive Statistics**: Provided an overview of key variables like Time Spent on Course and Quiz Scores.

**GLM Application**: Applied a logistic regression model to predict course completion based on engagement metrics.

**Model Selection**: Used AIC and BIC criteria to select the best-fit model.

**Model Assessment**: Evaluated model performance using predicted probabilities and ROC curve analysis.

**Time Spent on Course**: Significant positive predictor of course completion.

**Number of Videos Watched**: Also positively associated with course completion.

**Quiz Scores**: Higher quiz scores increase the likelihood of completing the course.

**Device Type**: Users on desktops are more likely to complete the course compared to mobile users.

**Interpretation of the ROC Curve:**

**AUC Value:**

Here, the auc_value is 0.8697, it indicates that the model has a good ability to distinguish between students who will complete the course and those who will not.

An AUC value of 0.8697 means that there is an 86.97% chance that the model will rank a randomly chosen student who completed the course higher than a randomly chosen student who did not.

**Conclusion:**

The ROC curve provides a visual assessment of the model's performance across different classification thresholds. The AUC quantifies this performance into a single value, making it easier to compare different models. A higher AUC value indicates a better-performing model, and in this case, it suggests that the logistic regression model is effective in predicting course completion.
